About p-adic stability

### 2014-12-03 examples/stability.tex
Paper reorganized and introduction (first version) added

\documentclass{sig-alternate}
%\documentclass{lms}
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amssymb}
\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}
\renewcommand*{\theHsection}{\thesection}

@@

\title{Tracking p-adic precision: examples}

@@

\begin{abstract}
\end{abstract}

@@

 \vspace{1mm}
 \noindent
 {\bf Keywords:} 
\medskip

@@

\section{Preliminaries}

In the next section, we will see that the derivative at some point $x$ 
of many standard operations, modelised by a function $f$, has a simple 
expression in term of $x$ and $f(x)$. In other terms, such functions $f$
satisfy:
\begin{equation}
\label{eq:diffequa}
f' = g \circ (f, \id)
\end{equation}
where $g$ is a given --- and hopefully rather simple --- function. 
The aim of these preliminaries is to study the differential equation
\eqref{eq:diffequa} (or, actually, a slight generalization of it) and
to derive from it some bounds on the growing function $\Lambda(f)$ 
defined in \cite{padicprec}.

\emph{In all this section, we assume that \textbf{the base field $K$ 
has characteristic $0$}.}

\begin{comment}

\subsection{Legendre transform of a compositum}

For $i \in \{1,2\}$, let $\varphi : \R \to \R \cap \{+\infty\}$ denote a 
convex function and $\varphi_i^\star$ be its Legendre transform defined 
by:
\begin{equation}
\label{eq:defLegendre}
\varphi_i^\star(u) = \sup_{v \in \R} \varphi(v) - uv.
\end{equation}
Set $\varphi = \varphi_1 \circ \varphi_2$. The aim of this subsection is 
to relate $\varphi^\star$ with $\varphi_1^\star$ and $\varphi_2^\star$.

\begin{lem}
\label{lem:Legendrecomp}
We assume that $\varphi_1$ and $\varphi_2$ are nondecreasing and that 
their slopes are all in $\N$. Then, for all nonnegative interger $s$:
$$\varphi^\star(s) =
\inf_{\ell, (n_i)} \varphi_1^\star(\ell) + \varphi_2^\star(n_1) + 
\cdots + \varphi_2^\star(n_\ell)$$
where the infimum is taken over all finite sequences of nonnegative
integers (of variable length) $n_1, \ldots, n_\ell$ with $n_1 + \cdots +
n_\ell = s$.
\end{lem}

\begin{proof}
Let $n_1, \ldots, n_\ell$ be a sequence as above. Taking the supremum 
in $v$ in the relation:
$$\varphi(v) - s v =
\varphi_1(\varphi_2(v)) - \ell \varphi_2(v) + \sum_{i=1}^\ell
(\varphi_2(v) - n_i v)$$
we get:
\begin{equation}
\label{eq:inegvarphi}
\varphi^\star(s) \leq \varphi_1^\star(\ell) + 
\varphi_2^\star(n_1) + \cdots + \varphi_2^\star(n_\ell).
\end{equation}
It remains then to prove that one can choose $\ell$ and $n_1, \ldots,
n_\ell$ such that the inequality above is actually an equality.

Let $x$ be the unique value such that the left (resp. right) derivative 
of $\varphi$ at $x$ is less than (resp. greater than) or equal to $s$. 
Let $\ell^-$ (resp. $\ell^+$) denote the left (resp. right) derivative 
of $\varphi_1$ at $\varphi_2(x)$ and $n^-$ (resp. $n^+$) denote the left 
(resp. right) derivative of $\varphi_2$ at $x$. On the one hand, $\ell^- 
n^- \leq s \leq \ell^+ n^+$ since $\ell^- n^-$ (resp. $\ell^+ n^+$) 
equals the left (resp. right) derivative of $\varphi$ at $x$). On the 
other hand:
$$\begin{array}{ll}
\varphi_1^\star(u) = \varphi(x) - u \varphi_2(x) & 
\text{for all } u \in [\ell^-,\, \ell^+] \\
\varphi_2^\star(u) = \varphi_2(x) - u x & 
\text{for all } u \in [n^-,\, n^+] \\
\varphi^\star(u) = \varphi(x) - u x & 
\text{for all } u \in [\ell^- n^-,\, \ell^+ n^+] \\
\end{array}$$
Therefore, the inequality \eqref{eq:inegvarphi} is an equality for all 
integer $\ell \in [\ell^-, \ell^+]$ and all sequences of integers 
$(n_i)_{1 \leq i \leq \ell}$ with $n_i \in [n^-, n^+]$ for all $i$ and 
$n_1 + \cdots + n_\ell = s$. It is then enough to prove that such a 
couple $(\ell, (n_i))$ exists but it is clear.
\end{proof}

\end{comment}

Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces 
$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$. Let 
$f : U \to V$ and $g : V \to W$ be two locally analytic functions around 
$0$ with $f(0) = 0$. The composition $h = g \circ f$ is then locally 
analytic around $0$ as well. Let us write the analytic expansion of $f$, 
$g$ and $h$ as follows:
$$f = \sum_{n \geq 0} f_n, \quad 
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n, \quad$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
aim of this subsection is to prove the following proposition.

\subsection{Bounding a growing function}
\label{ssec:boundLambdaf}

We now study a slight generalization of the differential equation 
\eqref{eq:diffequa}, which is:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g$ and $h$ are known locally analytic functions and $f$ is the 
unknown, which is assumed to be locally analytic as well. The function
$f$ goes from $U$ to $V$, two open subsets in $K$-Banach spaces $E$ 
and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
is again an open subset in a $K$-Banach space $G$. Consequently, $g$
goes from $V \times W$ to the vector space $\mathcal L(E,F)$ gathering
all continuous linear applications $E \to F$.
In what follows, we always assume that $V$ and $W$ contains the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
because (1)~we can always shift $f$ and $h$ (and $g$ accordingly) so 
that they both vanish at $0$ and (2)~in order to apply the theory of 
\cite{padicprec}, the derivative $f'(0)$ needs to be surjective and 
therefore \emph{a fortiori} nonzero.

Let us write the analytic expansion of $f$, $g$ and $h$ as follows: 
$$f = \sum_{n \geq 0} f_n, \quad
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
Following \cite{padicprec}, we define $\NP(f) : \R \to \R \cup 
\{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
\log \Vert f_n \Vert$ for all integer $n$ and denote by $\Lambda(f)$ its 
Legendre transform. We define $\NP(g)$, $\Lambda(g)$, $\NP(h)$ and
$\Lambda(h)$ similarly.

We consider a nonnegative real number $\alpha$ such that $\Vert n! \Vert 
\geq e^{-\alpha n}$ for all positive integer $n$. A suitable value for 
$\alpha$ is
$$\alpha = - \frac p {p-1} \cdot \log \Vert p \Vert > 0$$
where $p$ denotes the characteristic of the residue field and
where, by convention, the above expression equals $0$ if $p = 0$
(we recall that we have assumed that $K$ has characteristic $0$).

\begin{prop}
\label{prop:boundLambdaf}
We keep the above notations. If the couple $(a,b)$ satisfies:
\begin{equation}
\label{eq:condAB}
b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)
\end{equation}
then $b \geq \Lambda(f)(a - \alpha)$.
\end{prop}

\begin{rem}
A simpler proof is possible if $\Z \subset \O_K$, \emph{i.e.} if
one can choose $\alpha = 0$. Indeed, under this extra assumption,
one can prove using Eq.~\eqref{eq:normderivative} that:
$$\Lambda(f') \geq \Lambda(f) - \id.$$
Combining this with Lemma 3.7 of \cite{padicprec}, we find:
$$\Lambda(f) - \id \leq \Lambda(g) \circ \max(\Lambda(f), \Lambda(h))$$
from which we can easily derive Proposition \ref{prop:boundLambdaf}.
\end{rem}

We want now to rephrase Proposition \ref{prop:boundLambdaf} in a more 
workable way. We assume that we are given two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We assume in addition that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$]{-}\infty, \nu]$\footnote{We note that this assumption is fullfiled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
$$\begin{array}{ll}
\tau_\nu(x) = x & \text{if } x \leq \nu \\
\hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise}
\end{array}$$
and:
$$\Lambda_f(x) = 
  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)$$
where we recall that $\alpha$ satisfies $\Vert n! \Vert \geq e^{-\alpha 
n}$ for all $n$.

\begin{cor}
\label{cor:boundLambdaf}
We have $\Lambda(f) \leq \Lambda_f$.
\end{cor}

\begin{proof}
Given $a \in \R$ and $b = a + \Lambda_g \circ \Lambda_h(a)$, we have to 
prove that $\Lambda(f)(a-\alpha) \leq b$ provided that $b \leq \nu$. 
Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to check 
that such couples $(a,b)$ satisfy \eqref{eq:condAB}. Clearly:
$$b \geq a + \Lambda(g) \circ \Lambda(h)(a)$$
since $\Lambda_g \geq \Lambda(g)$, $\Lambda_h \geq \Lambda(h)$ and
$\Lambda_g$ is nondecreasing. Furthermore, from $b \leq \nu$, we get:
$$\Lambda_g(b) = \min_{x \in \R} \Lambda_g(x) \leq \Lambda_g \circ 
\Lambda_h(a).$$
Thus $a + \Lambda_g(b) \leq a + \Lambda_g \circ \Lambda_h(a) =
b$ and we are done.
\end{proof}

@@

Figure \ref{fig:area} illustrates Corollary \ref{cor:boundLambdaf}. The 
blue plain line represents the graph of the function $\Lambda_f$. A 
quick computation shows that, on a neighborhood of ${-}\infty$, this 
function is given by:
$$\Lambda_f(x) = x + \alpha + \mu$$
where $\mu$ is the value that $\Lambda_g$ takes on the interval 
$]{-}\infty, \nu]$. Corollary \ref{cor:boundLambdaf} says that the
graph of $\Lambda(f)$ lies below the plain blue line. We remark
moreover that the Taylor expansion of $f(x)$ starts with the term
$g(0) x$. Hence, on a neighborhood on ${-}\infty$, we have 
$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
get:
$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
  \quad \forall x \in \R.$$
In other words, the graph of $\Lambda(f)$ lies above the brown line.
Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
the dashed blue line defined as the line of slope $2$ passing through
the first break point of the blue plain line --- which has coordinate 
$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
must coincide with the brown line until it meets the dashed blue line 
and then has to stay in the green area.

As a consequence of the above discussion, we derive the following 
proposition which can be directly combined with Proposition 3.12 of 
\cite{padicprec}.

@@

\noindent
\textbf{About the choice of the function $h$}

\smallskip

At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
function $h$ in our differential equation. The aim of this was to be 
more flexible and possibly get this way better bounds on $\Lambda(f)$. 
In this paragraph, we discuss a recommended choice for this function 
$h$.

@@

@@

This phenomenon is general: if $P$ is a polynomial over $\Q_p$ known at 
some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
(\alpha)$ vanish at the given precision) can be computed at precision 
roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
where $c$ is some constant (depending on $P$ and $\alpha$ but not on
$N$).

\section{Multivariate polynomials}

@@

\section{Vector spaces}

@@

%\bibliographystyle{plain}
%\bibliography{roebib/Biblio,extras}

>>>>

\documentclass{sig-alternate}
\usepackage[utf8]{inputenc}

\usepackage{amsmath,amssymb}
\usepackage{amsrefs}
\usepackage[usenames,dvipsnames]{color}
\usepackage{stmaryrd}
\usepackage{enumerate}
\usepackage[algoruled,vlined,english,linesnumbered]{algorithm2e}
\usepackage[pdfpagelabels,colorlinks=true,citecolor=blue]{hyperref}
\usepackage{comment}
\usepackage{tikz}

@@

\title{About p-adic stability}

@@

\begin{abstract}
Based on the framework developed in \cite{caruso-roe-vaccon:14a}, we study the 
$p$-adic stability of many classical algorithms involving many usual
algebraic objects (matrices, univariate and multivariate polynomials,
vector spaces). We observe in particular that generic algorithms ---
which are designed to work over an arbitrary field --- are generally
very unstable.
\end{abstract}

@@

 \vspace{1mm}
 \noindent
 {\bf Keywords:} $p$-adic precision
\medskip

@@

For about twenty years, the use of $p$-adic methods in symbolic 
computation was gaining more and more popularity: they were succesfully 
used for instance 
to compute compose products of polynomials 
\cite{boston-gonzalez-perdry-schost:05a}, 
to produce hyperelliptic curves of genus $2$ with complex multiplication 
\cite{gaudry-houtmann-weng-ritzenthaler-kohel:06a},
to compute isogenies between elliptic curves \cite{lercier-sirvent:08a} 
and, last but not least,
to count points on certain varieties (mainly curves) using $p$-adic 
cohomologies (\emph{cf} \cite{kedlaya:01a,lauder:04a} and many
followers).
However, a general framework allowing a precise study of $p$-adic 
precision --- which is a main issue we encounter when trying to work 
with $p$-adic numbers --- was designed only recently in 
\cite{caruso-roe-vaccon:14a}.

The present paper, which may be considered as a continuation of 
\cite{caruso-roe-vaccon:14a}, aims at using the theory of \emph{loc. cit.} in 
order to analyse the $p$-adic stability of many standard algorithms
for computing with all basic algebraic structures that are numbers,
matrices, univariate and multivariate polynomials, series and vector
spaces. For each considered algorithm, we will follow the same protocol: 
first, we will compute the theoretical loss of precision of the 
underlying question using the framework of \cite{caruso-roe-vaccon:14a} and then 
compare it to those observed in available implementations in 
\textsc{pari} \cite{pari}, \textsc{magma} \cite{magma} and/or 
\textsc{sage} \cite{sage}.

The conclusion of our analysis is that generic algorithms which are 
designed to work over an arbitrary field are often quite unstable. Very 
roughly, denoting by $D$ the number of divisions performed by the 
algorithm under analysis and by $q$ the cardinality of the residue 
field, it appears that the numbers of lost digits is about $O(\frac D 
q)$ whereas the theoretical loss of precision is closer than $O(\log_q 
D)$ --- and sometimes even $O(1)$. We would like to underline in 
particular that the instability phenomema are much more apparent when 
the residue field is small. It is the reason why we will always take
$K = \Q_2$ is our examples.

\noindent
{\bf Organization of the paper.}
The section \ref{sec:theory} introduces the theory of ultrametric 
precision developped in \cite{caruso-roe-vaccon:14a} and completes it on a technical 
point. The analysis of standard algorithms are treated in the next
sections: \S\S \ref{sec:numbers}--\ref{sec:vectorspaces} are 
successively devoted to the case of numbers, matrices, univariate 
polynomials, multivariate polynomials and finally vector spaces.

\noindent
{\bf Notations.}
Throughout the paper, the letter $K$ will refer to a \emph{complete 
discrete valuation field of characteristic $0$}, that is either a finite 
extension of the field of $p$-adic numbers $\Qp$ or a field of Laurent 
series over a field of characteristic $0$. We denote by $\val : K \to \Z 
\cup \{+\infty\}$ of $K$ --- it is the $p$-adic valuation when $K = \Qp$ 
and the usual valuation of a Laurent series when $K = k((t))$ --- and by 
$\Vert \cdot \Vert$ the corresponding norm.

\section{The theory of p-adic precision}
\label{sec:theory}

\subsection{The main lemma and its consequences}

\todo{Recall results of \cite{caruso-roe-vaccon:14a}.}

We saw in the core of the article that the differential of an operation
encodes the intrinsic loss/gain of precision when performing this
operation. The aim of this appendix is to compute the differential of
many common operations on numbers, polynomials and matrices. Surprisingly,
we observe that all differentials we will consider are rather easy to
compute even if the underlying operation is quite involved (\emph{e.g.}
Gr\"obner basis).

In what follows, we use freely the ``method of physicists'' to compute
differentials: given a function $f$ differentiable at some point $x$, we
consider a small perturbation $dx$ of $x$ and write $f(x+dx) = y + dy$
by expanding LHS and neglecting terms of order $2$. The differential of
$f$ at $x$ is then the linear mapping $dx \mapsto dy$.

\subsection{A bound on a growing function}
\label{ssec:boundLambdaf}

In the next section, we will compute the derivative of many standard 
operations and remark that it often has a simple expression in term of 
the input and the output. In other terms, the function $f$ modeling
such an operation satisfies a differential equation of the form
$f' = g \circ (f, \id)$
where $g$ is a given --- and hopefully rather simple --- function. The 
aim of this subsection is to study this differential equation and to 
derive from it some bounds on the growing function $\Lambda(f)$.
We shall actually work with the slight more general differential 
equation:
\begin{equation}
\label{eq:diffequah}
f' = g \circ (f, h).
\end{equation}
Here $g$ and $h$ are known locally analytic functions and $f$ is the 
unknown, which is assumed to be locally analytic as well. The function
$f$ goes from $U$ to $V$, two open subsets in $K$-Banach spaces $E$ 
and $F$ respectively. The function $h$ goes from $U$ to $W$ where $W$
is again an open subset in a $K$-Banach space $G$. Consequently, $g$
goes from $V \times W$ to the vector space $\mathcal L(E,F)$ gathering
all continuous linear applications $E \to F$.
In what follows, we always assume that $V$ and $W$ contains the origin, 
$f(0) = 0$, $h(0) = 0$ and $g(0) \neq 0$. There assumptions are harmless 
for two reasons: first, we can always shift $f$ and $h$ (and $g$ 
accordingly) so that they both vanish at $0$ and second, in order to 
apply the theory of \cite{caruso-roe-vaccon:14a}, the derivative $f'(0)$ needs to be 
surjective and therefore \emph{a fortiori} nonzero.

We assume that we are given in addition two nondecreasing convex 
functions $\Lambda_g$ and $\Lambda_h$ such that $\Lambda(g) \leq 
\Lambda_g$ and $\Lambda(h) \leq \Lambda_h$. We assume in addition that 
there exists $\nu$ such that $\Lambda_g$ is constant on the interval 
$]{-}\infty, \nu]$\footnote{We note that this assumption is fullfiled if 
we take $\Lambda_g = \Lambda(g)$ because we have assumed that $g(0)$ 
does not vanish.}. We introduce the functions $\tau_\nu$ and $\Lambda_f$ 
defined by:
$$\begin{array}{rll}
&\tau_\nu(x) = x & \text{if } x \leq \nu \\
& \hphantom{\tau_\nu(x)} = {+} \infty & \text{otherwise} \medskip \\
\text{and} &
\multicolumn{2}{l}{\Lambda_f(x) = 
  \tau_\nu \circ (\id + \Lambda_g \circ \Lambda_h)(x + \alpha)}
\end{array}$$
where $\alpha$ is a real number satisfying $\Vert n! \Vert \geq 
e^{-\alpha n}$ for all $n$. A suitable value for $\alpha$ is 
$$\alpha = - \frac p {p-1} \cdot \log \Vert p \Vert > 0$$ 
where $p$ denotes the characteristic of the residue field and where, by 
convention, the above expression equals $0$ if $p = 0$ (we recall that 
we have assumed that $K$ has characteristic $0$).

\begin{prop}
\label{prop:boundLambdaf}\label{PROP:BOUNDLAMBDAF}
We have $\Lambda(f) \leq \Lambda_f$.
\end{prop}

\begin{proof}
See Appendix \ref{app:proof}.
\end{proof}

@@

Figure \ref{fig:area} illustrates Proposition \ref{prop:boundLambdaf}. 
The blue plain line represents the graph of the function $\Lambda_f$. A 
quick computation shows that, on a neighborhood of ${-}\infty$, this 
function is given by $\Lambda_f(x) = x + \alpha + \mu$
where $\mu$ is the value that $\Lambda_g$ takes on the interval 
$]{-}\infty, \nu]$. Proposition \ref{prop:boundLambdaf} says that the
graph of $\Lambda(f)$ lies below the plain blue line. We remark
moreover that the Taylor expansion of $f(x)$ starts with the term
$g(0) x$. Hence, on a neighborhood on ${-}\infty$, we have 
$\Lambda(f)(x) = x + \log \Vert g(0) \Vert$. Using convexity, we 
get:
$$\Lambda(f)(x) \geq x + \log \Vert g(0) \Vert, 
  \quad \forall x \in \R.$$
In other words, the graph of $\Lambda(f)$ lies above the brown line.
Furthermore, we know that the slopes of $\Lambda(f)$ are all integral
because $f$ is locally analytic. Hence, $\Lambda(f)$ cannot lie above
the dashed blue line defined as the line of slope $2$ passing through
the first break point of the blue plain line --- which has coordinate 
$(y_0 - \alpha - \mu, y_0)$ with $y_0 = \min(\Lambda_h^{-1}(\nu) + \mu, 
\nu)$. As a conclusion, we have proved that the graph of $\Lambda(f)$ 
must coincide with the brown line until it meets the dashed blue line 
and then has to stay in the green area.

As a consequence of the above discussion, we derive the following 
proposition which can be directly combined with Proposition 3.12 of 
\cite{caruso-roe-vaccon:14a}.

@@

\noindent
{\bf About the choice of the function $h$}.
At the beginning of \S \ref{ssec:boundLambdaf}, we have introduced a 
function $h$ in our differential equation. The aim of this was to be 
more flexible and possibly get this way better bounds on $\Lambda(f)$. 
In this paragraph, we discuss a recommended choice for it.

@@

\section{Numbers}
\label{sec:numbers}

The most basic operations are, of course, sums, products and inverses of
elements of $K$. Their differential are well-known and quite easy to
compute: if $z = x + y$ (resp. $z = xy$, resp $z = \frac 1 x$), we have
$dz = dx + dy$ (resp. $dz = x \cdot dy + dx \cdot y$, resp $dz = -
\frac{dx}{x^2}$).

Slightly more interesting is the $n$-th power map $f$ from $K$ to
itself. Its differential $f'(x)$ is found by differentiating the
equation $y = x^n$, obtaining $dy = n x^{n-1} dx$.
Hence $f'(x)$ maps the ball $B_K(r)$ to $B_K(\lvert nx^{n-1} \rvert \cdot r)$. According to Lemma \ref{lem:main}, this
means that the behavior of the precision depends on the absolute value
of $n$. By the ultrametric inequality, we always have $\lvert n \rvert \leq 1$ but
may have $\lvert n \rvert = 0$ if the characteristic of $K$ divides $n$ or
$\lvert n \rvert = p^{-k}$ if $K$ is $p$-adic with $\val(n) = k$.  In the first case,
$f'(x)$ also vanishes and Lemma \ref{lem:main} does
not apply.  In the second, Lemma \ref{lem:main} reflects the well-known fact that
raising a $p$-adic number to the $p^k$-th power increases the precision
by $k$ extra digits.

@@

This phenomenon is general: if $P$ is a polynomial over $\Qp$ known at 
some finite precision $O(p^N)$, a root $\alpha$ of $P$ having possibly 
multiplicity $m$ (\emph{i.e.} $P(\alpha), P'(\alpha), \ldots, P^{(m-1)} 
(\alpha)$ vanish at the given precision) can be computed at precision 
roughly $O(p^{N/m})$. However, if we know in advance that $\alpha$ has 
multiplicity $m$ then we can compute it at precision $O(p^{N-c})$ 
where $c$ is some constant (depending on $P$ and $\alpha$ but not on
$N$).

\section{Multivariate polynomials}
\label{sec:multivar}

@@

\section{Vector spaces}
\label{sec:vectorspaces}

@@

\appendix

\section{Proof of Proposition \ref{prop:boundLambdaf}}
\label{app:proof}

Let $U$, $V$ and $W$ be three open subsets is some $K$-Banach spaces 
$E$, $F$ and $G$ respectively. We assume that $0 \in U$, $0 \in V$. Let 
$f : U \to V$ and $g : V \to W$ be two locally analytic functions around 
$0$ with $f(0) = 0$. The composition $h = g \circ f$ is then locally 
analytic around $0$ as well. Let us write the analytic expansion of $f$, 
$g$ and $h$ as follows:
$$f = \sum_{n \geq 0} f_n, \quad 
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n, \quad$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively. The 
aim of this subsection is to prove the following intermediate result.

\subsection{Bounding a growing function}

We now go back to the setting of Proposition \ref{prop:boundLambdaf}.
We write the analytic expansion of $f$, $g$ and $h$ as follows: 
$$f = \sum_{n \geq 0} f_n, \quad
g = \sum_{n \geq 0} g_n, \quad
h = \sum_{n \geq 0} h_n$$
where $f_n$, $g_n$ and $h_n$ are the restrictions to the diagonal of 
some symmetric $n$-linear forms $F_n$, $G_n$ and $H_n$ respectively.
Following \cite{caruso-roe-vaccon:14a}, we define $\NP(f) : \R \to \R \cup 
\{+\infty\}$ as the greatest convex function such that $\NP(f)(n) \geq - 
\log \Vert f_n \Vert$ for all integer $n$ and denote by $\Lambda(f)$ its 
Legendre transform. We define $\NP(g)$, $\Lambda(g)$, $\NP(h)$ and
$\Lambda(h)$ similarly.
We recall that $\alpha$ is a nonnegative real number such that $\Vert n! 
\Vert \geq e^{-\alpha n}$ for all positive integer $n$.

\begin{lem}
\label{lem:boundLambdaf}
We keep the above notations. If the couple $(a,b)$ satisfies:
\begin{equation}
\label{eq:condAB}
b \geq a + \Lambda(g)\big( \! \max(b, \, \Lambda(h) (a)) \big)
\end{equation}
then $b \geq \Lambda(f)(a - \alpha)$.
\end{lem}

\begin{rem}
A simpler proof is possible if $\Z \subset \O_K$, \emph{i.e.} if
one can choose $\alpha = 0$. Indeed, under this extra assumption,
one can prove using Eq.~\eqref{eq:normderivative} that:
$$\Lambda(f') \geq \Lambda(f) - \id.$$
Combining this with Lemma 3.7 of \cite{caruso-roe-vaccon:14a}, we find:
$$\Lambda(f) - \id \leq \Lambda(g) \circ \max(\Lambda(f), \Lambda(h))$$
from which we can easily derive Lemma \ref{lem:boundLambdaf}.
\end{rem}

We can now conclude the proof of Proposition \ref{prop:boundLambdaf} as 
follows. Given $a \in \R$ and $b = a + \Lambda_g \circ \Lambda_h(a)$, we 
have to prove that $\Lambda(f)(a-\alpha) \leq b$ provided that $b \leq 
\nu$. Thanks to Proposition \ref{prop:boundLambdaf}, it is enough to 
check that such couples $(a,b)$ satisfy \eqref{eq:condAB}. Clearly:
$b \geq a + \Lambda(g) \circ \Lambda(h)(a)$
since $\Lambda_g \geq \Lambda(g)$, $\Lambda_h \geq \Lambda(h)$ and
$\Lambda_g$ is nondecreasing. Furthermore, from $b \leq \nu$, we get:
$$\Lambda_g(b) = \min_{x \in \R} \Lambda_g(x) \leq \Lambda_g \circ 
\Lambda_h(a).$$
Thus $a + \Lambda_g(b) \leq a + \Lambda_g \circ \Lambda_h(a) =
b$ and we are done.

\bibliographystyle{plain}
\bibliography{../roebib/Biblio,extras}
